---
title: "Poverty Rates"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE
# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE 
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))
# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)                     
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

The poverty rate reflects the percentage of the population in poverty. Data comes from the following ACS tables:

* S1701 (everything but age)
* B17001 (age)

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth County, NC', 'Guilford County, NC', 'Durham County, NC')
df <- read_csv('data/poverty_all_years_counties.csv') %>%
  # only keep selected NC counties, the state of NC, and US data
  filter(geo_description %in% geo_areas,
         # only need rows showing the percent below poverty level, don't need count data
         str_detect(description, '^Percent'))
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall employment rate of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
df %>%
  # only keep total poverty rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$')) %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Poverty rate by geographic area',
         color = 'Comparison Unit',
         y = 'Poverty rate (%)') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1)) +
    theme_minimal()
```

#### Interpretation

* In 2017, Forsyth County had the highest poverty rate among comparison geographic units. But, the difference between Forsyth and Durham counties is not statistically significant.
* Forsyth County's poverty rates have consistently dropped since 2012. The differences in poverty rates between 2017 and 2012, 2013, and 2014 are statistically significant. 

#### Data

Below is the raw data for poverty rates by geographies.

```{r include = to_include}
total_data <- df %>%
  # only keep total poverty rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$')) %>%
  select(geo_description, year, estimate, moe, se, cv)
# create dataset formatted for Tableau
tableau_data <- total_data %>%
  # add type of column
  mutate(type = 'Comparison Community',
         subtype = 'Total Population')
# create table
ff_data_dt(total_data, c('Geographic area', 'Year', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`.

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)
df %>%
  # only keep total employment rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$'),
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total employment rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$'),
         # only keep Forsyth
         geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Poverty rate by ethnicity

#### Plot

The plot below shows poverty rates by ethnicity and year.

```{r}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description)
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    #facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth County poverty rates by ethnicity',
         color = 'Ethnicity',
         y = 'Poverty rate (%)') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=ethnicity)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Poverty rate by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Poverty rate (%)',
       x = '') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
```

#### Interpretation

* The African American poverty rate has fallen almost 10 percentage points since 2012.
* The White, non-Hispanic poverty rate has remained stable over the past 10 years. Over this time period, no years have statistically significant differences from 2017.
* Most of the Hispanic/Latinx values have high CV scores (over 12).  Therefore, our confidence in Hispanic/Latinx estimates is weak.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
ethnicity_data <- ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv)
# add to dataset formatted for Tableau
tableau_data <- ethnicity_data %>%
  # add type of column
  mutate(type = 'Race and Ethnicity',
         subtype = ethnicity) %>%
  # delete ethnicity column since this is now in the subtype column
  select(-ethnicity) %>%
  bind_rows(tableau_data, .)
  
ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)
ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year','geo_description', 'ethnicity'))
```

The next table has ethnicity z-scores for all years within Forsyth County.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'ethnicity'))
```

------

### Poverty by age

Age groups are in table B17001. The block below imports and cleans data from this table.


```{r}
# create list of age descriptions that will be created
age_descriptions <- c('Under 5 years',
                      '5 to 17 years',
                      '18 to 34 years',
                      '35 to 64 years',
                      '65 years and over')
age <- read_csv('data/poverty_age_groups.csv') %>%
  # only keep selected NC counties, the state of NC, and US data
  filter(geo_description %in% geo_areas) %>%
  # extract age group and place age group name in its own columns
  mutate(age = str_extract(description, '- [0-9].*|- Under.*'),
         # remove '- ' from age group
         age = str_replace(age, '- ', ''),
         # exract whether row signifies above or below poverty and place in its own column
         level = str_extract(description, 'below poverty|above poverty')) %>%
  # filter out NA values for age group
  filter(!is.na(age)) %>%
  # create new age bins
  mutate(age = str_replace_all(age, '^5 y.*|^6 t.*|^12.*|^15.*|^16.*', age_descriptions[2])) %>%
  mutate(age = str_replace_all(age, '^18.*|^25.*', age_descriptions[3])) %>%
  mutate(age = str_replace_all(age, '^35.*|^45.*|^55.*', age_descriptions[4])) %>%
  mutate(age = str_replace_all(age, '^65.*|^75.*', age_descriptions[5])) %>%    
  # we don't need information about sex, so group by year, above/below poverty, and age group and sum
  group_by(geo_description, year, level, age) %>%
  # calculate aggregate counts and aggregate count MOE for aggregate of males and females
  summarize(estimate = sum(estimate),
            moe = sqrt( sum(moe^2) ))
    
# create separate dataframe for the total number of people in each age group
# calculate by adding number of people below poverty line to those above poverty line
age_total <- age %>%
  # group by year and age group
  group_by(geo_description, year, age) %>%
  # calculate aggregate counts and aggregate count MOE for each age group
  summarize(estimate = sum(estimate),
            moe = sqrt( sum(moe^2) ))
# filter out those above poverty because we only need percentage below poverty
age <- age %>%
  filter(level == 'below poverty') %>%
  # add age group totals to data frame; needed for creating percentages
  left_join(age_total, by = c('geo_description', 'year', 'age'), suffix = c('_below', '_total'))
# calculate percentages, along with MOE, se, and cv
age <- ff_acs_perc(age$estimate_below, age$moe_below,
                        age$estimate_total, age$moe_total) %>%
  # combine with dataset used to create percentages
  bind_cols(age, .) %>%
  ungroup() %>%
  # only keep needed years
  select(geo_description, year, age, prop:cv) %>%
  rename(estimate = prop) %>%
  # multiply percentages by 100, so their format is the same as the format for other indicators
  mutate(estimate = estimate * 100)
```

#### Plot

*The 5 to 17 year age group does not contain data for all years. This group was created by merging together other age groups, including one group of just five year olds. ACS does not show data for five year olds in 2008 due to the small sample size, and without data for this group the 5 to 17 year old group could not be created.*

```{r}
age %>%
    filter(geo_description =='Forsyth County, NC') %>%
    ggplot(aes(year, estimate, color = age)) +
      geom_line(size = 1) +
      labs(title = 'Forsyth County poverty rate by age group',
           color = 'Age group',
           y = 'Poverty rate (%)') +
      scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                         minor_breaks = NULL) +
      # change the order of age groups in legend so they appear youngest to oldest
      scale_color_discrete(breaks=age_descriptions) +
      theme_minimal()
```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Poverty rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
         y = 'Poverty rate (%)',
         x = '') +
    # change the order of age groups in legend so they appear youngest to oldest
    scale_fill_discrete(breaks=age_descriptions) +
    theme_minimal() +
    theme(axis.text.x=element_text(angle=45,hjust=1))
```

#### Interpretation

* Among all comparison geographies, Forsyth County had the highest poverty rate for children under five. But, only the US and Forsyth County difference is statistically significant.
* Within Forsyth County, children under five experienced a 13% drop in poverty between 2013 and 2014. This drop was statistically significant, and the difference between 2013 and 2017 is also statistically significant.

#### Data

Age data is below.

```{r include = to_include}
age_data <- age %>%
  select(geo_description, year, age, estimate, moe, se, cv)
#  add to dataset for Tableau
tableau_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-age) %>%
    bind_rows(tableau_data, .)
  
ff_data_dt(age_data,
           c('Geographic area', 'Year', 'Age group', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)
age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'age'))
```

Here are the age comparisons for all years within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'age'))
```

### Tableau dataset

This is the final dataset that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneded columns
  select(year, geo_description, type, subtype, estimate) %>%
  # removes rows containing demographic data for comparison communities 
  # in all years except the latest year
  ff_remove_comparisons() %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```